ML Estimation of Matrix--$t$ Distributions with Expectation Maximization

### 2017-01-22 paper.tex
Finished editing the EM section.

\subsection{EM formulas for matrix--$t$ distribution}

@@

Utilizing the equalities \eqref{eq:intderivative1} and \eqref{eq:intderivative2} with adequate rearrangements we obtain the closed form expression of $Q_j(\theta,\theta_k)$ while ignoring the irrelevant constant terms
\begin{equation}
\begin{split}Q_j(\theta,\theta_k) \sim \nu\left\{\log|W_j^{(k)}|+\psi_{p}\left(\frac{m+\nu_k+p-1}{2}\right)\right\}-2\log\Gamma_{p}\left(\frac{\nu+p-1}{2}\right)-p\log|\Omega|\\
-(m+\nu_k+p-1)\textrm{tr}\left[W_j^{(k)}\left\{\Sigma+(T_j-M)\Omega^{-1}(T_j-M)^{\textrm{t}}\right\}\right]+(\nu+p-1)\log |\Sigma|,
\end{split}
\label{eq:fsimp}
\end{equation}
where the matrices $W_j^{(k)}$ are defined as
\begin{equation}
W_j^{(k)}=\left\{\Sigma_k+(T_j-M_k)\Omega_k^{-1}(T_j-M_k)^{\textrm{t}}\right\}^{-1}.
\end{equation}

To maximize $Q(\theta,\theta_k)$ with respect to $\theta$ we compute the gradient of each parameter and equate them to $0$ to obtain $\theta_{k+1}$. For parameter $M$ we have
\begin{equation}
\sum_{j\in J}W_j^{(k)}\left(T_j-M\right)\Omega^{-1}=0.
\end{equation}
As $\Omega$ is nonsingular we obtain $M_{k+1}$ as an average of the input data weighted by $W_j^{(k)}$
\begin{equation}
M_{k+1}=\left(\sum_{j\in J}W_j^{(k)}\right)^{-1}\sum_{j\in J}W_j^{(k)}T_j.
\end{equation}
Maximizing $Q(\theta,\theta_k)$ with respect to $\Sigma$ we obtain $\Sigma_{k+1}$ in terms of the inverse of the mean of $W_j^{{k}}$
\begin{equation}
\Sigma_{k+1}=\frac{\nu+p-1}{m+\nu_k+p-1}\left(\frac{1}{|J|}\sum_{j\in J}W_j^{(k)}\right)^{-1}. 
\end{equation}
Note that $\Sigma_{k+1}$ given above has dependence on $\nu$. Similarly for $\Omega$ we obtain
\begin{equation}
\Omega_{k+1}=\frac{m+\nu_k+p-1}{p}\left(\frac{1}{|J|}\sum_{j\in J}(T_j-M_{k+1})^{\textrm{t}}W_j^{(k)}(T_j-M_{k+1})\right).
\end{equation}

For $\nu$ in the case when $\Sigma$ is a free parameter to be estimated we have
\begin{equation}
p\log\left(\frac{n_{t}+p-1}{m+n+p-1}\right)+\frac{1}{N}\sum_{i}^{N}\log|W_{i}|-\log\left|\frac{1}{N}\sum_{i}^{N}W_{i}\right|+\psi_{p}\left(\frac{m+n+p-1}{2}\right)-\psi_{p}\left(\frac{n_{t}+p-1}{2}\right)=0
\end{equation}
otherwise if $\Sigma_{t}=\Sigma$ is fixed and does not depend on
$n_{t}$:
\begin{equation}
\log|\Sigma_{t}|-\log\det\left(\frac{1}{N}\sum_{i}^{N}W_{i}\right)+\psi_{p}(\frac{1}{2}(m+n+p-1))-\psi_{p}(\frac{1}{2}(n_{t}+p-1))=0
\end{equation}

\section{Implementation}\label{sec:implement}

>>>>

\subsection{EM formulae for matrix--$t$ distribution}

@@

Utilizing the equalities \eqref{eq:intderivative1} and \eqref{eq:intderivative2} with adequate rearrangements we obtain the closed form expression of $Q_j(\theta,\theta_k)$ while ignoring the irrelevant constant terms
\begin{equation}
\begin{split}Q_j(\theta,\theta_k) \sim \nu\left\{\log|W_j^{(k)}|+\psi_{p}\left(\frac{m+\nu_k+p-1}{2}\right)\right\}-2\log\Gamma_{p}\left(\frac{\nu+p-1}{2}\right)-p\log|\Omega|\\
-(m+\nu_k+p-1)\textrm{tr}\left[W_j^{(k)}\left\{\Sigma+(T_j-M)\Omega^{-1}(T_j-M)^{\textrm{t}}\right\}\right]+(\nu+p-1)\log |\Sigma|,
\end{split}
\label{eq:fsimp}
\end{equation}
where the matrices $W_j^{(k)}$ are defined as
\begin{equation}
W_j^{(k)}:=\left\{\Sigma_k+(T_j-M_k)\Omega_k^{-1}(T_j-M_k)^{\textrm{t}}\right\}^{-1}.
\end{equation}

To maximize $Q(\theta,\theta_k)$ with respect to $\theta$ we compute the gradient of each parameter and equate them to $0$ to obtain $\theta_{k+1}$. For parameter $M$ we have
\begin{equation}
\sum_{j\in J}W_j^{(k)}\left(T_j-M\right)\Omega^{-1}=0.
\end{equation}
As $\Omega$ is nonsingular we obtain $M_{k+1}$ as an average of the input data weighted by $W_j^{(k)}$
\begin{equation}
M_{k+1}=\left(\sum_{j\in J}W_j^{(k)}\right)^{-1}\sum_{j\in J}W_j^{(k)}T_j.
\end{equation}
Maximizing $Q(\theta,\theta_k)$ with respect to $\Omega$ we obtain $\Omega_{k+1}$
\begin{equation}
\Omega_{k+1}=\frac{m+\nu_k+p-1}{p}\left(\frac{1}{|J|}\sum_{j\in J}(T_j-M_{k+1})^{\textrm{t}}W_j^{(k)}(T_j-M_{k+1})\right).
\end{equation}
For the rest of the parameter(s), we consider two possible situations.
\begin{itemize}
\item $\Sigma$ and $\nu$ are unknown: In this situation, maximizing $Q(\theta,\theta_k)$ with respect to $\Sigma$ we obtain $\Sigma_{k+1}$ in terms of the inverse of the mean of $W_j^{{(k)}}$
\begin{equation}
\Sigma_{k+1}(\nu)=\frac{\nu+p-1}{m+\nu_k+p-1}\left(\bar{W}^{(k)}\right)^{-1},\quad \bar{W}^{(k)}:=\frac{1}{|J|}\sum_{j\in J}W_j^{(k)}\label{eq:sigma}
\end{equation}
$\Sigma_{k+1}$ has linear dependence on $\nu$. Due to this, we substitute $\Sigma_{k+1}$ back into $Q(\theta,\theta_k)$ and maximize it with repect to $\nu$. We remark that
\[\tilde{Q}(\nu):=Q\left(\left\{M_{k+1},\Sigma_{k+1}(\nu),\Omega_{k+1},\nu\right\},\theta_k\right)\]
is a concave function of $\nu$ for all $\nu>0$. The maximum only attains at positive $\nu$ if $\tilde{Q}'(0)$ is positive. For this case we obtain $\nu_{k+1}$ by solving the equation $\tilde{Q}'(\nu)=0$, which has the form
\begin{equation}
h(\nu+p-1)-h(m+\nu_k+p-1)+\frac{1}{|J|}\sum_{j\in J}\log|W_j^{(k)}|-\log\left|\bar{W}^{(k)}\right|=0
\end{equation}
with $h(x):=p\log(x)-\psi_p(x/2)$.
\item $\Sigma$ is known and $\nu$ is unknown: In this situation, $Q\left(\left\{M_{k+1},\Sigma,\Omega_{k+1},\nu\right\},\theta_k\right)$ is also a concave function of $\nu$. The gradient equation of $\nu$ is given by
\begin{equation}
\log|\Sigma|-\log\left|\bar{W}^{(k)}\right|+\psi_{p}\left(\frac{m+\nu_k+p-1}{2}\right)-\psi_{p}\left(\frac{\nu+p-1}{2}\right)=0.
\end{equation}
\end{itemize}

\section{Implementation}\label{sec:implement}
In the actual implementation of
